---
title: "Get started with adoptr"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quickstart}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r setup-knitr, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 5
)
```



## Two-stage designs for clinical trials

In prospectively planned confirmatory clinical trials there has been a 
long-standing debate about the merits of allowing pre-planned interim analyses
to stop or modify an ongoing trial, cf. 
[BAUER *et al.*, 2016](https://onlinelibrary.wiley.com/doi/pdf/10.1002/sim.6472).
It is common consensus that such two-stage designs should still maintain 
strict control of the maximal type one error rate despite the data dependent
decision at the interim analysis.
Most authors use combination function approaches to combine either the
independent p values or test statistics from the respective stages.
Doing so with pre-specified weights maintains the type one error rate control.

With **adoptr** we provide tools to explore optimal two-stage designs for
one or two-arm clinical trials with (asymptotically) normally distributed
and stage-wise independent test statistics.
In [Pilz *et al.*, 2019](https://doi.org/10.1002/sim.8291),
it was argued that such techniques are not necessary, in fact,
that they are counter-productive, since the pre-specified weights might be
sub-optimal.
Instead, one may see the problem of defining an optimal two-stage designs as
a variational problem to find the optimal configuration 
$$
\mathcal{D}:=\big(n_1, c_1^e, c_1^f, n_2(\cdot), c_2(\cdot)\big)
$$ 
directly for a given objective function and potentially a set of constraints.
The relaxation of this mixed-integer problem to the real numbers can then
be solved with standard methods from analysis.
E.g., one could construct an implicit solution by solving the corresponding
Euler-Lagrange equation for given first-stage and then find an optimal
first-stage configuration, 
cf. [Pilz *et al.*, 2019](https://doi.org/10.1002/sim.8291).
Instead, the **adoptr** package implements a direct numerical solution 
algorithm by embedding the variation problem via a set of pivot points in 
the $\mathbb{R}^k$ and solving the corresponding finite-dimensional problem 
using gradient-free numerical optimization procedures (COBYLA / nloptr).


# Example

Consider the two-arm case comparing two group means mean.
Let $\delta$ be the difference in means and assume that the aim is to 
reject the null hypothesis $\mathcal{H}_0: \delta \leq 0$. 
Further assume that a power of 80% at $\delta=0.4$ and a type-one-error rate
of 2.5% on the boundary of the null hypothesis $\delta\leq0.0$ is required.

In this situation, the one-sided t-test would require a sample size of
$51$ individuals per-group.
If we were willing to accept the fact that the sample size of a two-stage
design is inherently unknown at the planning stage, 
we could reduce the *expected* sample size under the alternative by optimizing
a two-stage design with respect to this quantity.
Since the two-stage design, by definition, has more degrees of freedom than
the single-stage design, it must be at least as good in terms of expected sample
size as the single-stage design - for which the expected sample size is 
always 51.


## Define hypotheses and data-generating mechanism

```{r load-adoptr}
library(adoptr)
```

First, we define the null and alternative hypotheses as prior distributions 
over the unknown parameter $\delta$.

```{r define-hypotheses}
H_0 <- PointMassPrior(.0, 1)
H_1 <- PointMassPrior(.4, 1)
```

Next, we define the data-generating mechanism (data distribution).
Currently, normally distributed test statistics with known variance,
and the normal approximation of binary endpoints are supported.
In this example, we choose a normal distribution, i.e., assume a continuous
outcome.
For an example with binomial data distribution, cf. the vignette
[composite scores](https://kkmann.github.io/adoptr/articles/composite-scores.html).

```{r define data-gernerating-mechanism}
datadist <- Normal(two_armed = FALSE)
```


## Specify optimization problem

We begin by defining the expected sample size under the alternative hypothesis:

```{r define-ess}
ess <- ExpectedSampleSize(datadist, H_1)
```

Similarly, we can define the power and the type-one-error rate by
integrating conditional power with respect to the alternative or 
null hypothesis respectively:

```{r define-power-toer}
power <- Power(datadist, H_1)
toer  <- Power(datadist, H_0)
```

To complete the problem specification, we only need to provide an initial 
design and an integration order for the Gaussian quadrature rule to be used for
the numerical integration of the variational objective between $c_1^f$ and 
$c_1^e$.
**adoptr** provides a heuristic method to obtain an initial design.
It is based on a one-stage design under specific constraints on type one error
rate and power.
However, it may be more efficient to use a situation-specific initial design.
```{r create-design}
initial_design <- get_initial_design(
  theta = .4,
  alpha = .025,
  beta  = .2,
  type  = "two-stage",
  dist  = datadist,
  order = 7L
)
```

The decision boundaries are heuristically picked in concordance with those of
the single-stage design.
We can then check whether the initial design fulfills the constraints on the 
operating characteristics:

```{r check-power}
evaluate(power, initial_design)
```

```{r check-toer}
evaluate(toer, initial_design)
```

We can now proceed with defining and solving the optimization problem:

```{r optimize}
opt_res <- minimize(
  
  ess,
  
  subject_to(
    power >= 0.8,
    toer  <= .025
  ),
  
  initial_design
)
```

Here, the default setting for the optimizer and box constraints seem to have 
worked out.
The numerical procedure converged after 

```{r nloptr-output, eval = FALSE}
opt_res$nloptr_return$iterations
```

iterations.

We can plot the resulting optimized design may again be plotted together with
its conditional power.

```{r plot-optimal-design}
plot(
  opt_res$design, 
  "Conditional power" = ConditionalPower(datadist, H_1)
)
```

Numerical integration yields an expected sample size of 

```{r}
evaluate(ess, opt_res$design)
```

which is way below the single-stage 51. 
We can check this results against simulated values 
```{r}
df_sim <- simulate(
  opt_res$design, 
  nsim = 10^6, 
  dist = datadist, 
  theta = .4, 
  seed = 42
)

n <- df_sim$n1 + df_sim$n2
mean(n)
sd(n) / sqrt(length(n))
```

The values from simulation and numerical integration are almost identical 
and the constraints are also met almost perfectly:
```{r}
evaluate(toer, opt_res$design)
evaluate(power, opt_res$design)
```



## Variations

The purpose of **adoptr** is to provide a high-level interface for exploring 
and comparing different designs and to experiment with different 
combinations of objective functions and constraints.
In particular, **adoptr** supports:

* Important predefined scores as (conditional) power or sample size functions as 
well as a framework for constraints on scores 
(cf. [conditional scores](https://kkmann.github.io/adoptr/articles/conditional-scores.html)).

* Arbitrary continuous or discrete 
[prior distributions](https://kkmann.github.io/adoptr/articles/working-with-priors.html) 
for the single location  parameter $\theta$ of the test statistic.

* Support for simple arithmetic operations on scores to quickly define weighted
sums etc. (cf. [composite scores](https://kkmann.github.io/adoptr/articles/composite-scores.html)).

* Relatively flexible framework for
[defining new scores](https://kkmann.github.io/adoptr/articles/defining-new-scores.html).
