---
title: "Working with priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 5
)
```



## Working with priors in **adoptr**

Internally, **adoptr** is built around the joint distribution of a test 
statistic and the unknown location parameter of interest given a sample size,
i.e.
$$
\mathcal{L}\big[(X_i, \theta)\,|\,n_i\big]
$$
where $X_i$ is the stage-$i$ test statistic and $n_i$ the corresponding sample
size.
The distribution class for $X_i$ is defined by specifying a `DataDistribution`
object, e.g., a normal distribution
```{r setup}
library(adoptr)

datadist <- Normal()
```
To completely specify the marginal distribution of $X_i$, the distribution of
$\theta$ must also be specified. 
The classical case where $\theta$ is considered fixed, emerges as special
case when a single parameter value has probability mass 1.



### Discrete priors

The simplest supported prior class are discrete `PointMassPrior` priors.
To specify a discrete prior, one simply specifies the vector of pivot points 
with positive mass and the vector of corresponding probability masses.
E.g., consider an example where the point $\delta = 0.1$ has probability mass 
$0.4$ and the point $\delta = 0.25$ has mass $1 - 0.4 = 0.6$. 
```{r discrete-prior}
disc_prior <- PointMassPrior(c(0.1, 0.25), c(0.4, 0.6))
```
For details on the provided methods, see `?DiscretePrior`.


### Continuous priors

**adoptr** also supports arbitrary continuous priors with support on compact 
intervals.
For instance, we could consider a prior based on a truncated normal via:
```{r}
cont_prior <- ContinuousPrior(
  pdf     = function(x) dnorm(x, mean = 0.3, sd = 0.2), 
  support = c(-2, 3)
)
```

For details on the provided methods, see `?ContinuousPrior`.


### Conditioning

In practice, the most important operation will be conditioning.
This is important to implement type one and type two error rate constraints.
Consider, e.g., the case of power.
Typically, a power constraint is imposed on a single point in the alternative,
e.g. using the constraint
```{r}
expected(ConditionalPower(
  Normal(), 
  PointMassPrior(.4, 1)
)) >= 0.8
```
If uncertainty about the true response rate should be incorporated in the design,
it makes sense to assume a continuous prior on $\theta$.
In this case, the prior should be conditioned for the power constraint to
avoid integrating over the null hypothesis:
```{r}
expected(ConditionalPower(
  Normal(), 
  condition(cont_prior, c(0, 3))
)) >= 0.8
```
